---
title: "COMP 4441 Final Project"
author: "Heather Lemon & Katie Chen"
date: "5/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse)
library(dplyr)
library(ggplot2)
```

# The question we would like to answer 
Do you think the car wreck fatality rate went up or down from 2019-2020? 
Did Covid have an impact on car wreck fatalities? 


#load data
```{r}
person2020 = read.csv("./person2020.csv")
dat20=data.frame(person2020)


person2019 = read.csv("./person2019.csv")
dat19=data.frame(person2019)

```

# extract entities 
```{r}
# total car wreck deaths in 2019&2020
table_death_count2020=table(dat20$DOA)

total2020=table_death_count2020[2]+table_death_count2020[3]

table_death_count2019=table(dat19$DOA)
total2019=table_death_count2019[2]+table_death_count2019[3]

df2019=data.frame(total2019)
df2020=data.frame(total2020)

counts= c(df2019$total2019, df2020$total2020)
years = c('2019', '2020')

df=data.frame(years, counts)

ggplot(data=df, aes(x=years, y=counts))+geom_bar(stat = "identity", width = .3, alpha=0.7, color="black")+ggtitle("Fatality Count per Year")

```
# Breaking down deaths per Month
```{r}

n19=table(dat19$MONTHNAME)
df19=data.frame(n19)

n20=table(dat20$MONTHNAME)
df20=data.frame(n20)

sort_by_month<-factor(df19, levels = month.name)

df_month_19=data.frame(sort_by_month)
df_month_20=data.frame(sort_by_month)

ggplot(data = df19, aes(x=Var1, y=Freq))+geom_point(stat = "identity")+geom_line()+ggtitle("DEATH COUNT PER MONTH 2019")+scale_x_discrete(limits=month.name)
ggplot(data = df20, aes(x=Var1, y=Freq))+geom_point(stat = "identity")+geom_line()+ggtitle("DEATH COUNT PER MONTH 2020")+scale_x_discrete(limits=month.name)

combined_df=rbind(df19, df20)

ggplot(data = combined_df, aes(x=Var1, y=Freq))+geom_point(data = df19, stat="identity") + geom_point(data = df20, stat="identity")+ggtitle("DEATH COUNT COMBINED")+scale_x_discrete(limits=month.name)

ggplot(data = df19, aes(x=Var1, y=Freq))+geom_point()+geom_line()+ggtitle("DEATH COUNT COMBINED")+scale_x_discrete(limits=month.name)
```
# apply filters for unknown data 
```{r}
age_19<-data.frame(AGE=dat19$AGE[dat19$AGE<115])
ggplot(data=age_19, aes(x=AGE))+geom_bar(width=.5)+ggtitle("AGE 2019")

age_f<-data.frame(AGE=dat20$AGE[dat20$AGE<115])
ggplot(data=age_f, aes(x=AGE))+geom_bar(width=.5)+ggtitle("AGE 2020")
```


#Speeding Violations
```{r}
v2020=read.csv("./violatn2019.CSV")
v2019=read.csv("./violatn2020.CSV")


#speeding codes
viol_df2019=data.frame(table(v2019$VIOLATION))
speeding2019=viol_df2019$Freq[20]+viol_df2019$Freq[21]+viol_df2019$Freq[22]+viol_df2019$Freq[23]+viol_df2019$Freq[24]+viol_df2019$Freq[25]

viol_df2020=data.frame(table(v2020$MVIOLATN))
speeding2020=viol_df2020$Freq[20]+viol_df2020$Freq[21]+viol_df2020$Freq[22]+viol_df2020$Freq[23]+viol_df2020$Freq[24]+viol_df2020$Freq[25]

speeding_counts= c(speeding2019, speeding2020)
years = c('2019', '2020')

df=data.frame(years, speeding_counts)

ggplot(data=df, aes(x=years, y=speeding_counts))+geom_bar(stat = "identity", width = .3)+ggtitle("Speeding Count per Year")

```
```
I would be hesitant to say that speeding was a major factor of why the increase in falities despite me having friends/family say that
they noticed an increase of people speeding. 
I found this article mentioning speeding as a reason for more deaths: https://www.thelongofirm.com/posts/colorado-car-accident-fatalities-increase-2020
```


#Accidents related to Alcohol 
```{r}
alc_df2019=data.frame(table(dat19$DRINKING))
alc_df2020=data.frame(table(dat20$DRINKING))

counts= c(alc_df2019$Freq[2], alc_df2020$Freq[2])
years = c('2019', '2020')

df_alc=data.frame(years, counts)

ggplot(data = df_alc, aes(x=years, y=counts))+geom_bar(stat = "identity", width = .3)+ggtitle("Alcohol Related Death Count")
```


# COVID case datasets
```{r}

#This section gets the total count of covid-19 cases from January-December 2020 
dat <- read.csv('national-history.csv')
#dat

#we care about aggregate positive 
#the column 'positive' from the last day of each month
#from January 2020 - December 2020
#then plot

#convert date to datetime
dat %>%
   mutate(date=as.Date(date))

#cut to all entries between January 2020 - December 2020
y_2020 <- filter(dat, dat$date <= '2020-12-31' )
y_2020

#take the final case counts from each month
# creates an identifier with which to group
y_2020$mon_yr <- format(as.Date(y_2020$date), '%Y-%m') 
#groups by created month identifier and then keeps only those rows with last(max) date
cases <- y_2020 %>% group_by(mon_yr) %>% filter(date == max(date))
cases

```



```{r}
#graph the positive cases

cases$logpositive = log(cases$positive)
cases$monthnames <- months(as.Date(cases$date))
cases$month<-factor(cases$monthnames, levels = month.name)

ggplot(data = cases, aes(x = month, y = logpositive)) +
  geom_point() +
  labs(x = "Month",
    y = "Postive Cases (Log)",
    title = "Positive Covid Cases in 2020 (Log form)") +   theme(axis.text.x=element_text(angle=45,hjust=1))

ggplot(data = cases, aes(x = month, y = positive)) +
  geom_point() +
  labs(x = "Month",
    y = "Postive Cases",
    title = "Positive Covid Cases in 2020") + theme(axis.text.x=element_text(angle=45,hjust=1))
```


```{r}
#covid cases as a bar chart

cases$logpositive = log(cases$positive)
cases$monthnames <- months(as.Date(cases$date))
cases$month<-factor(cases$monthnames, levels = month.name)

ggplot(cases, aes(month)) + geom_bar(aes(weight=positive), position="dodge", fill= 'lightblue') + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(x = "Month",
    y = "Postive Cases",
    title = "Positive Covid Cases in 2020")

ggplot(cases, aes(month)) + geom_bar(aes(weight=logpositive), position="dodge", fill= 'lightblue') + theme(axis.text.x=element_text(angle=45,hjust=1)) +  labs(x = "Month",
    y = "Postive Cases (log)",
    title = "Positive Covid Cases in 2020 in Log Format")
```

#this is the part where we compare covid-19 data from 2020 to road accident data in 2020

```{r}
# total car wrecks in 2020
car_wrecks <- read.csv('person2020.csv')

```

```{r}
#counts of car wrecks by month
df <- data.frame(aggregate(car_wrecks, by=list(car_wrecks$MONTHNAME), FUN=length))
names(df) <- c('Month','Crashes')

#stick it in a new data frame
columns <- c('Month','Crashes')
data2020 <- df[,columns]

data2020

```



```{r}
#combine covid 2020 with car crashes 2020
covid <- data.frame(cases$monthnames, cases$positive)
names(covid) <- c("Month", "Cases")

total <- merge(data2020, covid, by.x = "Month", by.y="Month")
names(total) <- c("Month", "Crashes", "Cases") #naming them so we don't throw an error

total$logCases <- log(total$Cases)
total$logCrashes<-log(total$Crashes)
total
```


```{r}
#graph the total 

# Reshape data frame
df_reshaped <- data.frame(x = total$Month,                           
                       y = c(total$logCases, total$logCrashes),
                       group = c(rep("Cases", nrow(df)),
                                 rep("Crashes", nrow(df))))

df_reshaped$x <- factor(df_reshaped$x, levels = month.name)

ggplot(df_reshaped, aes(x, y, col = group)) +  geom_line(aes(group = group)) + theme(axis.text.x=element_text(angle=45,hjust=1)) +  labs(x = "Month",
    y = "Crashes and Cases",
    title = "Road Crashes and Covid Cases by Month in 2020 (Log)")

```

#Ridge Regression Method (Math)

Ridge Regression is used to fit a model that has multicollinearity while trying to 
minimize the (RSS) sum of square residuals. 
Col-linearity in regression is the event of two or more variables being highly linearly related. 


In terms of relationships between variables, our use case refers to the target or dependent 
variable as the number of car wrecks. The independent variable or predictor variable could be any of these; 
speeding, weather, age, covid, alcohol, drugs, etc. 


#The Ridge Regession Esimator 
$$\hat \beta \left( \lambda \right) = \left( X^TX+\lambda I_p \cdot_p \right)^ \minus1 $$







