---
title: "COMP 4441 Final Project"
author: "Heather Lemon & Katie Chen"
date: "2022/05/23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse)
library(dplyr)
library(ggplot2)
library(reshape2)
library(lubridate)
library(glmnet)
```

# The questions we would like to answer 
Do you think the car wreck fatality rate went up or down from 2019-2020? 
Did Covid have an impact on car wreck fatalities? 


#Load people dataset
```{r}
person2020 = read.csv("./person2020.csv")
dat20=data.frame(person2020)


person2019 = read.csv("./person2019.csv")
dat19=data.frame(person2019)

```

# Total Car Wreck Deaths in 2020&2019 
```{r}
table_death_count2020=table(dat20$DOA)

total2020=table_death_count2020[2]+table_death_count2020[3]

table_death_count2019=table(dat19$DOA)
total2019=table_death_count2019[2]+table_death_count2019[3]

df2019=data.frame(total2019)
df2020=data.frame(total2020)

counts= c(df2019$total2019, df2020$total2020)
years = c('2019', '2020')

df=data.frame(years, counts)

ggplot(data=df, aes(x=years, y=counts))+geom_bar(stat = "identity", width = .3, alpha=0.7, color="black")+ggtitle("Fatality Count per Year")

percent_increase = ((counts[2]-counts[1])/counts[1]) * 100
percent_increase
```
# Breaking down deaths per Month
```{r}

n19=table(dat19$MONTHNAME)
df19=data.frame(n19)

n20=table(dat20$MONTHNAME)
df20=data.frame(n20)

sort_by_month<-factor(df19, levels = month.name)

df_month_19=data.frame(sort_by_month)
df_month_20=data.frame(sort_by_month)

ggplot(data = df19, aes(x=Var1, y=Freq))+geom_point(stat = "identity")+geom_line()+ggtitle("DEATH COUNT PER MONTH 2019")+scale_x_discrete(limits=month.name)
ggplot(data = df20, aes(x=Var1, y=Freq))+geom_point(stat = "identity")+geom_line()+ggtitle("DEATH COUNT PER MONTH 2020")+scale_x_discrete(limits=month.name)

combined_df=rbind(df19, df20)

ggplot(data = combined_df, aes(x=Var1, y=Freq))+geom_point(data = df19, stat="identity")+geom_line() + geom_point(data = df20, stat="identity")+ggtitle("DEATH COUNT COMBINED")+scale_x_discrete(limits=month.name)
```
# Group by Age 
```{r}
age_19<-data.frame(AGE=dat19$AGE[dat19$AGE<115])
ggplot(data=age_19, aes(x=AGE))+geom_bar(width=.5)+ggtitle("AGE 2019")

age_f<-data.frame(AGE=dat20$AGE[dat20$AGE<115])
ggplot(data=age_f, aes(x=AGE))+geom_bar(width=.5)+ggtitle("AGE 2020")
```


# Speeding Violations
```{r}
v2020=read.csv("./violatn2019.CSV")
v2019=read.csv("./violatn2020.CSV")


#speeding codes
viol_df2019=data.frame(table(v2019$VIOLATION))
speeding2019=viol_df2019$Freq[20]+viol_df2019$Freq[21]+viol_df2019$Freq[22]+viol_df2019$Freq[23]+viol_df2019$Freq[24]+viol_df2019$Freq[25]

viol_df2020=data.frame(table(v2020$MVIOLATN))
speeding2020=viol_df2020$Freq[20]+viol_df2020$Freq[21]+viol_df2020$Freq[22]+viol_df2020$Freq[23]+viol_df2020$Freq[24]+viol_df2020$Freq[25]

speeding_counts= c(speeding2019, speeding2020)
years = c('2019', '2020')

df=data.frame(years, speeding_counts)

ggplot(data=df, aes(x=years, y=speeding_counts))+geom_bar(stat = "identity", width = .3)+ggtitle("Speeding Count per Year")

percent_speeding = ((speeding_counts[2]-speeding_counts[1])/speeding_counts[1]) * 100
percent_speeding
```
```
I would be hesitant to say that speeding was a major factor of why the increase in falities despite me having friends/family say that
they noticed an increase of people speeding. 
I found this article mentioning speeding as a reason for more deaths: https://www.thelongofirm.com/posts/colorado-car-accident-fatalities-increase-2020
```


# Alcohol 
```{r}
alc_df2019=data.frame(table(dat19$DRINKING))
alc_df2020=data.frame(table(dat20$DRINKING))

counts= c(alc_df2019$Freq[2], alc_df2020$Freq[2])
years = c('2019', '2020')

df_alc=data.frame(years, counts)

ggplot(data = df_alc, aes(x=years, y=counts))+geom_bar(stat = "identity", width = .3)+ggtitle("Alcohol Related Death Count")

percent_alcohol = ((counts[2]-counts[1])/counts[1]) * 100
percent_alcohol 
```


#Ridge Regression Method (Math)

Ridge Regression is used to fit a model that has multicollinearity while trying to 
minimize the (RSS) sum of square residuals. 
Col-linearity in regression is the event of two or more variables being highly linearly related. 


In terms of relationships between variables, our use case refers to the target or dependent 
variable as the number of car wrecks. The independent variable or predictor variable could be any of these; 
speeding, weather, age, covid cases, alcohol, drugs, etc. 

Also note that ridge regression requires the data to be standardized such that each predictor variable has a mean of 0 and a standard deviation of 1.

*least squares regression tries to find coefficient estimates that minimize the sum of squared residuals (RSS) (https://www.statology.org/ridge-regression-in-r/)

As the tuning parameter gets larger the value of the coefficients shrink towards 0. 

#The Ridge Regession Esimator 

$$ \sum [yi- ŷ_i^2] + λ \sum_{k=1}^{k} y_i^2 $$

Another form in linear algebra. 
$$\hat{\beta}\left(\lambda\right)=\left(X^TX+\lambda I_{p \times p}\right)^{-1}X^TY$$ 

Linear algebra plays a significant role in understanding the lambda value as well as the rank and matrix operations. 
Selecting a valid lambda value is crucial to getting the best fit for the model.  

$$y=\ f\left(x_1,\ x_2,\ x_3...\ x_n\right) $$

To put it simply we want to know which predictor $$(x_i)$$ had the greatest affect on the dependent variable y.


# Ridge Regression Analysis 

# Cleaning and Reformatting Data

The Covid-19 dataset taken from the Covid Tracking Project was reformatted to include the final positive case counts from each month between January and December 2020. Given that the original data set tracked cumulative daily counts, this was calculated by taking the total case counts on the last day of each month.

```{r}

#This section gets the total count of covid-19 cases from January-December 2020 
dat <- read.csv('national-history.csv')

#convert date to datetime
dat<-dat %>%
   mutate(date=as.Date(date))

#cut to all entries between January 2020 - December 2020
y_2020 <- filter(dat, dat$date <= '2020-12-31' )

#take the final case counts from each month
# creates an identifier with which to group
y_2020$mon_yr <- format(as.Date(y_2020$date), '%Y-%m') 
#groups by created month identifier and then keeps only those rows with last(max) date
cases <- y_2020 %>% group_by(mon_yr) %>% filter(date == max(date))
cases

```

# Graphing and Displaying Reformatted Data:

In graphing the number of positive covid-19 cases, a log scale was added to better show the overall number of cases nationwide. 

```{r}
#graph the positive cases

cases$logpositive = log(cases$positive)
cases$monthnames <- months(as.Date(cases$date))
cases$month<-factor(cases$monthnames, levels = month.name)

ggplot(data = cases, aes(x = month, y = logpositive)) +
  geom_point() +
  labs(x = "Month",
    y = "Postive Cases (Log)",
    title = "Positive Covid Cases in 2020 (Log form)") +   theme(axis.text.x=element_text(angle=45,hjust=1))

ggplot(data = cases, aes(x = month, y = positive)) +
  geom_point() + 
  labs(x = "Month",
    y = "Postive Cases",
    title = "Positive Covid Cases in 2020") + theme(axis.text.x=element_text(angle=45,hjust=1)) 
```

For additional comparison, a bar chart was added to reflect the total covid-19 case counts in both the normal and log scales.

```{r}
#as a bar chart

cases$logpositive = log(cases$positive)
cases$monthnames <- months(as.Date(cases$date))
cases$month<-factor(cases$monthnames, levels = month.name)

ggplot(cases, aes(month)) + geom_bar(aes(weight=positive), position="dodge", fill= 'lightblue') + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(x = "Month",
    y = "Postive Cases",
    title = "Positive Covid Cases in 2020")

ggplot(cases, aes(month)) + geom_bar(aes(weight=logpositive), position="dodge", fill= 'lightblue') + theme(axis.text.x=element_text(angle=45,hjust=1)) +  labs(x = "Month",
    y = "Postive Cases (log)",
    title = "Positive Covid Cases in 2020 in Log Format")
```

# Cleaning and Reformatting Road Incident Data:

Road accident data taken from FARS (Fatality Analysis Reporting System) 2020 was cleaned and reformatted to reflect aggregated data by month. The reformatted data was then merged with covid-19 data to show the following:

- month of the year
- total number of road incidents reported by FARS 
- total positive covid-19 case counts 
- the average age of individuals involved in FARS incidents 
- total number of drinking violations
- total number of speeding violations

The following rmd blocks show the complete process of reformatting and merging the two data sets:

# 1. Read in FARS 2020 data set

```{r}
# total car wrecks in 2020
car_wrecks <- read.csv('person2020.csv')

```

# 2. Aggregate the total number of entries per month 

```{r}
#counts of car wrecks by month
df <- data.frame(aggregate(car_wrecks, by=list(car_wrecks$MONTHNAME), FUN=length))
names(df) <- c('Month','Crashes')

#stick it in a new data frame
columns <- c('Month','Crashes')
data2020 <- df[,columns]

data2020

```

# 3. Merge (inner join) reformatted covid-19 data with FARS entry counts on month

Log scales were added for both covid-19 cases and crash data for scalable comparison.

```{r}
#combine covid 2020 with car crashes 2020
covid <- data.frame(cases$monthnames, cases$positive)
names(covid) <- c("Month", "Cases")

total <- merge(data2020, covid, by.x = "Month", by.y="Month")
names(total) <- c("Month", "Crashes", "Cases") #naming them so we don't throw an error

total$logCases <- log(total$Cases)
total$logCrashes<-log(total$Crashes)
total
```

# 4. Age was taken from the original FARS 2020 data set and aggregated by calculating the average age per month of individuals involved in each FARS incident. Entries reflecting unknown ages '999' were dropped.

```{r}

#Since our data set is split by month, we can get
#average age each month

car_wrecks.age<-filter(car_wrecks, AGE != 999)
#car_wrecks.age$AGE

car_wrecks.age <- car_wrecks.age %>%
  group_by(MONTHNAME) %>%
  summarise(result = mean(AGE) )

names(car_wrecks.age) <- c("Month", "Avg_Age")

car_wrecks.age


```

# 5. Cleaned and reformatted age data was merged.

```{r}
#join the age table on total
total.a <- merge(total, car_wrecks.age, by.x = "Month", by.y="Month")
total.a

```

#6. Drinking reports were calculated based on known reported drinking incidents, where the DRINKING column in the original FARS data set = 1. A sum for each month was calculated and inserted into a dataframe. 

```{r}
#here we aggregate Drinking reports
car_wrecks.drink<-filter(car_wrecks, DRINKING==1)
#car_wrecks.drink

car_wrecks.drink <- car_wrecks.drink %>%
  group_by(MONTHNAME) %>%
  summarise(result = sum(DRINKING) )

names(car_wrecks.drink) <- c("Month", "drink_counts")

car_wrecks.drink

```

# 7. Drinking counts were merged into the data frame. 

```{r}
#merge drinking with total

total.b <- merge(total.a, car_wrecks.drink, by.x = "Month", by.y="Month")
total.b

```

#8. Speeding related incidents were manually taken from the FARS annual report. 
The following vector reflects the total speeding violation counts per month in 2020. A month vector was added and speeding related incidents transformed into a data frame. Dataframe total.c reflects the final data frame transformation with all our predictor variables. 

```{r}
#add speeding-related incidents
#manually taken from page 19
#Table 4. Monthly Traffic Fatalities, by Speeding Involvement, Alcohol-Impaired Driving, and Passenger Vehicle Occupant Restraint Use, 2019 and 2020

speeding <- c(733, 718, 735, 761, 1007, 1189, 1151, 1089, 1079, 999, 883, 914)
month<-month.name

car_wrecks.speeding <- data.frame(month, speeding)
names(car_wrecks.speeding)<-c("Month", "Speeding")

#merge into the main df

total.c <- merge(total.b, car_wrecks.speeding, by.x = "Month", by.y="Month")
total.c

```

The following graph shows the final monthly aggregates for each predictor in our data set. 

```{r}
#graph the totals
total.c$logAge <- log(total.c$Avg_Age)
total.c$logDrinks <- log(total.c$drink_counts)
total.c$logSpeed <- log(total.c$Speeding)

# Reshape data frame
df_reshaped <- data.frame(x = total.c$Month,                           
                       y = c(total.c$logCases, total.c$logCrashes, total.c$logAge, total.c$logDrinks, total.c$logSpeed),
                       group = c(rep("Covid Cases", nrow(df)),
                                 rep("Crashes", nrow(df)),
                                 rep("Average Age", nrow(df)),
                                 rep("Drinking Violations", nrow(df)),
                                 rep("Speeding Violations", nrow(df))
                                 ))

df_reshaped$x <- factor(df_reshaped$x, levels = month.name)

ggplot(df_reshaped, aes(x, y, col = group)) +  geom_line(aes(group = group)) + geom_point(aes(group = group)) +  theme(axis.text.x=element_text(angle=45,hjust=1)) +  labs(x = "Month",
    y = "Predictors",
    title = "Counts of Predictor Variables on Car Accidents by Month in 2020 (Log Scale)")

```

# Ridge Regression Analysis:

The ridge regression was performed using the final transformed data frame "total.c". In this analysis, the response variable is defined as the total number of incidents per month reported by FARS. Columns 'Cases', 'Avg_Age', 'drink_counts', and 'Speeding' were formatted into an 'x' matrix. Using the glmnet package, a ridge regression model was fitted onto the x and y variables, and an optimal lambda value minimizing the mean squared error was chosen.

```{r}
#begin ridge regression

#define response variable and predictors
y <- total.c$Crashes
x<-data.matrix(total.c[,c('Cases','Avg_Age','drink_counts', 'Speeding')])

#fit ridge regression model
model <- glmnet(x, y, alpha = 0)
#summary(model)

#perform k-fold cross-validation to find optimal lambda value
cv_model <- cv.glmnet(x, y, alpha = 0)

#find optimal lambda value that minimizes test MSE
best_lambda <- cv_model$lambda.min
best_lambda

#produce plot of test MSE by lambda value
plot(cv_model) 

```

Based on the values of the coefficients of the best model, we can conclude that the number of positive covod-19 cases had very little to no effect on the overall number of FARS driving incidents in 2020. In fact it appears that the average age of drivers had the largest impact, followed by overall alcohol consumption and speeding incidents.

```{r}

#coefficients of the best model
best_model <- glmnet(x, y, alpha = 0, lambda = best_lambda)
coef(best_model)

```

The model below shows the Ridge trace plot for each of the coefficients as the log lambda increases. As lambda increases, the coefficients trend towards 0. 

```{r}
#produce Ridge trace plot
plot(model, xvar = "lambda")

```

# Verifying our Model

To verify the accuracy of our model, we predicted some y-values using our best model and used these values to calculate the total sum of squares (sst) and the sum of squared errors (sse). From these values, we know that the R^2 value of our model is about 0.94. Thus, our model explains about 94% of the variation in our data.

```{r}
#use fitted best model to make predictions
y_predicted <- predict(model, s = best_lambda, newx = x)

#find SST and SSE
sst <- sum((y - mean(y))^2)
sse <- sum((y_predicted - y)^2)

#find R-Squared
rsq <- 1 - sse/sst
rsq
```



